---
title: Quality control and preprocessing for Framingham data
author: Kenny Westerman
output: pdf_document
---

```{r prereqs, include=FALSE}
lapply(c("knitr","tidyverse","minfi","wateRmelon","RefFreeEWAS","FlowSorted.Blood.450k"),
       library, character.only=T)
opts_chunk$set(warning=FALSE, message=FALSE, cache.path="../cache/meth_data_prep/", tidy=TRUE, tidy.opts=list(width.cutoff=30))
```

```{r load-data, cache=TRUE}
dataDir <- "../data/"  # At the moment, this only contains a few samples for development
targets <- read.metharray.sheet(base=dataDir, pattern="^sample_sheet4.csv$") %>%
  dplyr::filter(Basename!="character(0)")
rgSet <- read.metharray.exp(targets=targets, extended=T)
```
```{r signal-plots, eval=FALSE}
# Boxplots to check on signals across samples -- not performed for many samples.
boxplot(log(getGreen(rgSet)))
boxplot(log(getRed(rgSet)))
```

QC: based on detection p-values, remove questionable probes and samples, each defined by having a certain percentage of undetected elements
```{r QC, cache=TRUE, dependson="load-data"}
mSet.filt <- pfilter(mn=rgSet, pnthresh=0.05, perc=5, pthresh=5)  # Samples and probes must have <5% of representatives with detection P>0.05
```

Normalization is performed using beta-mixture quantile normalization (BMIQ; Teschendorff et al. 2013).
```{r normalization, cache=TRUE, dependson="QC"}
mSet.filt.bmiq_matrix <- BMIQ(mSet.filt)
mSet.filt.bmiq <- RatioSet(Beta=mSet.filt.bmiq_matrix)
```

Calculate cell counts using the reference-based approach from Houseman et al. 2012.
```{r cell-counts, cache=TRUE, dependson="normalization"}
data(FlowSorted.Blood.450k.compTable)
cellRef <- as.matrix(FlowSorted.Blood.450k.compTable)[,c('CD8T','CD4T','NK','Bcell','Mono','Gran')]  # Only needs cols representing actual cell types
commonCpGs <- intersect(rownames(mSet.filt.bmiq), rownames(cellRef))  # Which CpG sites are common to my set and the cell type reference?
estCellCounts <- projectMix(Y=mSet.filt.bmiq_matrix[commonCpGs,], Xmat=cellRef[commonCpGs,])
boxplot(estCellCounts, main="Calculated cell type proportions")
```

Create final table with technical, and cell count covariates.
```{r covariate-table, cache=TRUE, dependson=c("load-data","cell-counts")}
targets_withKey <- mutate(targets, Sample_pKey=paste(Slide, Sample_Well, sep="_"))
estCellCounts_df <- rownames_to_column(as.data.frame(estCellCounts), var="Sample_pKey")
meth_covData <- inner_join(targets_withKey, estCellCounts_df, by="Sample_pKey")
```


```{r save-data, cache=TRUE, dependson="covariate-table"}
save("mSet.filt.bmiq", file="../data/preprocessed_methylation_data.Rdata")
save("meth_covData", file="../data/methylation_covariates.Rdata")
```


